---
title: "Swirl_Graphics_Devices_in_R"
author: "Mr. Sachin B."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1. Graphics Device?
<p>It is something where you can make a plot appear, either a **screen device**, such as a **window** on your computer, OR a **file device**. There are several different kinds of **file devices** with particular characteristics and hence uses. These include **PDF, PNG, JPEG, SVG, and TIFF**.</p>

<p> When you make a plot in R, it has to be "sent" to a specific graphics device.

- Usually this is the screen (the default device), especially when you're doing exploratory work.
- You'll send your plots to files when you're ready to publish a report, make a presentation, or send info to colleagues.
</p>
<br>

####  How you access your screen device depends on what computer system you're using.

- On a **Mac** the screen device is launched with the call `quartz()`
- On **Windows** you use the call `windows()`
- On **Unix/Linux** `x11()`.

- On a given platform (**Mac, Windows, Unix/Linux**) there is only one screen device, and obviously not all graphics devices are available on all platforms (i.e. **you cannot launch 'windows()' on a 'Mac'**).

To see what graphics devices are available on your system.
```{r eval=FALSE}
?Devices
```


### 2. Two basic approaches to plotting.

#### 2.1. The first, **plotting to the screen**, is the most common.

- It's simple - you call a plotting function like `plot()`, `xyplot()`, or `qplot()` (which you call depends on the plotting system you favor, so that the plot appears on the screen.
- Then you **annotate** (add to) the plot if necessary.

```{r}
# Create Plot
with(faithful,plot(eruptions,waiting))

# Annotate Plot
title(main = "Old Faithful Geyser data")

# Show you the current plotting device, the screen.
dev.cur()
```
<br>

#### 2.2. The second way to create a **plot is to send it to a file device**.

- Depending on the type of plot you're making, you explicitly launch a graphics device, e.g., a **pdf file**.
- Type the command `pdf(file="myplot.pdf")` to launch the file device.
- This will create the pdf file **myplot.pdf** in your working directory.

```{r}
# Create the pdf file myplot.pdf in your working directory
pdf(file="myplot.pdf")

# Call the plotting function (if you are using a file device, no plot will appear on the screen).
with(faithful,plot(eruptions,waiting))

# Annotate Plot
title(main = "Old Faithful Geyser data")

# Finally, when plotting to a file device, you have to close the device with the command

dev.off() #This is very important!
```

### 3. Two basic types of file devices: 1. vector 2. bitmap devices

- These use different formats and have different characteristics.
    1. **Vector formats** are good for line drawings and plots with solid colors using a modest number of points.
    2. **bitmap formats** are good for plots with a large number of points, natural scenes or web-based plots.
    
#### 3.1. vector formats
We'll mention 4 specific vector formats.

1. The first is **pdf**
    - This is useful for line-type graphics and papers.
    - It resizes well, is usually portable, but it is not efficient if a plot has many objects/points.
    
2. The second is **svg** 
    - which is XML-based, scalable vector graphics.
    - This supports animation and interactivity and is potentially useful for web-based plots.

3. **win.metafile**
    - a Windows-only metafile format
    
4. **postscript** (ps)
    - an older format which also resizes well, is usually portable, and can be used to create encapsulated postscript files. 
    - Unfortunately, Windows systems often don’t have a postscript viewer.
    
#### 3.2. bitmap formats
We'll mention 4 specific vector formats.

1. The first is **png** (Portable Network Graphics)
    - which is good for line drawings or images with solid colors.
    - It uses lossless compression (like the old GIF format), and most web browsers can read this format natively.
    - In addition, png is good for plots with many points, but it does not resize well.

2. In contrast, **jpeg** files
    - jpeg files are good for photographs or natural scenes.
    - They use lossy compression, so they're good for plots with many points.
    - Files in jpeg format don't resize well, but they can be read by almost any computer and any web browser.
    - They're not great for line drawings.

3.  **tiff**
    - an older lossless compression meta-format.

4. **bmp**
    - which is a native Windows bitmapped format.

### 4. dev.cur() and dev.off()

Although it is possible to open multiple graphics devices (screen, file, or both), when viewing multiple plots at once, **plotting can only occur on one graphics device at a time**.

The currently active graphics device can be found by calling
```{r}
dev.cur()
```

Adding new graphic device to see changes
```{r}

# Adding new graphic device
pdf(file="test.pdf")

# Now checking current device
dev.cur()
```

- As you might have guessed, every open graphics device is assigned an integer greater than or equal to 2.
- You can change the active graphics device with dev.set(<integer>) where <integer> is the number associated with the graphics device you want to switch to.
```{r}
dev.set(2)

dev.cur()
```

To close the device
```{r}
dev.off()
```

Now rerun dev.cur() to see what integer your plotting window is assigned.
```{r}
dev.cur()
```

### 5. copy a plot from one device to another

This can save you some time but beware! Copying a plot is not an exact operation, so the result may not be identical to the original. R provides some functions to help you do this. 
<br>
The function **dev.copy** copies a plot from one device to another, and dev.copy2pdf specifically copies a plot to a PDF file.
```{r}

# Call the plotting function
with(faithful,plot(eruptions,waiting))

# Annotate Plot
title(main = "Old Faithful Geyser data")

# Finally, copy a plot from one device (screen device) to another (file device)
dev.copy(png,file = "geyserplot.png")

# This is very important!
dev.off() 
```