---
title: "Graphic Devices in R"
author: "Mr. Sachin B."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. What is Graphic Device?

A graphics device is something where you can make a plot appear

- A window on your computer (screen device)
- A PDF file (file device)
- A PNG or JPEG file (file device)
- A scalable vector graphics (SVG) file (file device)

When you make a plot in R, it has to be "sent" to a specific graphics device.

The most common place for a plot to be "sent" is the screen device

- Screen device on Mac is launched with the quartz()
- Screen device on is launched with windows()
- Screen device on Unix/Linux is launched with x11()

### 2. How to create Plot
I. For quick visualizations, usually the **screen device is used**.
II. For plots that printed out or incorporated into a document such as papers/reports, slide presentations, usually a **file device is more appropriate**.
  
#### I. 
  1. call a plotting Functions like 
    
- `plot` in **base**, 
- `xyplot` in **lattice**, or 
- `qplot` in **ggplot2** will default to sending a plot to the screen device


```{r}

# 100 random numbers using  normal distribution 
x<-rnorm(100)

# Histogram
hist(x)

```