---
title: "R_Graphs"
author: "Mr. Sachin B."
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Plotting Systems
There are three different plotting systems in R and they each have different characteristics and modes of operation.

They three systems are:

1. Base Plotting System
2. Lattice System
3. ggplot2 System
  
#### 1.1. Base Plotting System
* It is oldest system in R and come with original version of R.
* Conceptual model for building plot is 'Artist's Palette' model. which means 'Start with blank canvas and built up from there'
  * First, Create box with some points in it.
  * then, add labels and then add regression line.
  * then, title, axis ticks etc.
* To do this, 'Start with `plot()` function' that generate plot.
* Then there are other functions that so called, Annotate the plot.
  * eg: text, lines, labels etc.
 
```{r}
 data(cars)
 
 ## Create the plot / draw canvas
 with(cars, plot(speed, dist))
 
 ## Add annotation
 title("Speed vs. Stopping distance")
 
```

#### 1.2. Lattice System
* The lattice plotting system is implemented in the lattice package which comes with every installation of R (although it is not loaded by default).
* Instead of piecing a plot together one by one through a series of commands, every plot is constructed with a series of commands, every plot is constructed with a **Single Function Call**
 * eg: `xyplot()`, `bwplot()` etc.
 * then, add labels and then add regression line.
 * then, title, axis ticks etc.
 
```{r}
library(lattice)

state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4, 1))
 
```

#### 1.3. ggplot2 System
* This comes from the grammar of graphics which lays out a set of principles for a kind of plotting and creates a kind of language or grammar for describing  different aspects of a plot.
* The ggplot2 system is implemented in the ggplot2 package, which is available from CRAN (it does not come with R).
* It split the difference between base and lattice system so it mixes ideas from both.
  * One hand, You can build plot incrementally by adding things one by one just like *base system*.
  * On the other hand, Lot of aesthetic calculations are done automatically without you having to directly control such as 'spacing' & 'labels' are putting in the right place just like *lattice system*
 * eg: `qplot()` function in *ggplot2*.
 * then, add labels and then add regression line.
 * then, title, axis ticks etc.
 
```{r}
library(ggplot2)

data(mpg)
qplot(displ, hwy, data = mpg)
 
```

### 2. graphics and grDevices Packages

The core plotting and graphic engine in R is encapsulated in the following packages:

1. graphics: Contains plotting functions for the 'base' graphics system including `plot()`,`hist()`,`boxplot()` and many more

2. grDevices: Contains all the code implementing the various graphics devices including `windows()`,`pdf()`,`png()` etc.


### 3. graphics package

### 3.1. Histogram
```{r}

# 100 random numbers using  normal distribution 
x<-rnorm(100)

# Histogram
hist(x)

```

### 3.2. Scatter Plot
```{r}

# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot
plot(x,y)

# Scatter Plot Practice
a <- rnorm(100)
b <- rnorm(100)
plot(a,b)

```

### 3.3. Plot Parameter

#### 3.3.1 `Margin`
```{r}

# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot Before Applying Margin (Default Margin)
plot(x,y)
```

```{r}
# Adding Margin to Scatter Plot
par(mar = c(2,2,2,2))

# Scatter Plot After Applying Margin
plot(x,y)

```


#### 3.3.2 `pch`
```{r}

# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)


# Scatter Plot using pch=20
plot(x,y,pch=20)

# Scatter Plot using pch=15
plot(x,y,pch=15)

# Scatter Plot using pch=10
plot(x,y,pch=10)

# Scatter Plot using pch=5
plot(x,y,pch=5)
```

#### To know more about pch symbol
```{r eval=FALSE }
example("points")
```

#### 3.3.3 `title`

##### Approach 1
```{r}
# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot
plot(x,y)

# Adding Title to Plot
title("My Scatterplot 1")

```

##### Approach 2
```{r}
# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot along with title
plot(x,y, main="Scatterplot 2")

```

#### 3.3.4 `text`
```{r}
# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot
plot(x,y,pch=8)

# Adding Text in a Plot
text(-2,-1,"MyText")

```

#### 3.3.5 `legend`
```{r}
# 100 random numbers using  normal distribution 
x <- rnorm(100)
y <- rnorm(100)

# Scatter Plot
plot(x,y,pch=8)

# Adding Legend to a Plot
legend("topleft",legend = "Data", pch = 8)

```

#### 3.3.6 `abline`
`abline` function adds one or more straight lines through the current plot.<br>

The basic syntax is of abline() is as follows:<br>

`abline(a=NULL, b=NULL, h=NULL, v=NULL, …)`

- a, b: single values that specify the intercept`(a)` and slope`(b)` of the line
- h: the y-value for the horizontal line
- v: the x-value for the vertical line
```{r}
#define dataset

df <- data.frame(x = c(1,1,2,3,4,4,5,6,7,7,8,9,10,10,11),
                 y = c(12,13,16,11,22,24,25,24,24,27,30,32,35,38,40))

#plot x and y values in dataset
plot(df$x, df$y, pch = 15)
```

##### 3.3.6.1 `abline`:`Horizontal Line`

```{r}
# To add a horizontal line at the value y = 25
plot(df$x, df$y, pch = 15)

abline(h = 25, col = 'blue', lwd = 2)
```

The following code illustrates how to add a horizontal solid line at the mean value of y along with two horizontal dashed lines at one standard deviation above and below the mean value


- Note that lwd = 2 specifies that we want the line width to be equal to 2 (default = 1).

- Note that lty = 2 specifies that we want the line to be dashed.

```{r}
#create scatterplot for x and y
plot(df$x, df$y, pch = 18)

#create horizontal line at mean value of y
abline(h = mean(df$y), col='blue', lwd = 2)

#create horizontal lines at one standard deviation above and below the mean value
abline(h = mean(df$y) + sd(df$y), col = 'red', lwd = 3, lty = 2)
abline(h = mean(df$y) - sd(df$y), col = 'red', lwd = 3, lty = 2)

```

##### 3.3.6.2 `abline`:`Vertical Line`
```{r}
# To add a vertical line at the value x = 6
plot(df$x, df$y, pch = 15)

abline(v = 6, col = 'blue', lwd = 2)
```

The following code illustrates how to add a vertical line at the mean value on a histogram

```{r}

#make this example reproducible
set.seed(0)

#create dataset with 1000 random values normally distributed with mean = 10, sd = 2
data <- rnorm(1000, mean = 10, sd = 2)

#create histogram of data values
hist(data, col = 'steelblue')

#draw a vertical dashed line at the mean value
abline(v = mean(data), lwd = 3, lty = 2)

```

##### 3.3.6.3 `abline`:`Regression Line`

The basic code to add a simple linear regression line to a plot in R is: `abline(reg_model)`

- where `reg_model` is a fitted regression line created by using the `lm()` function.

The following code illustrates how to add a fitted linear regression line to a scatterplot

```{r}
#create scatterplot of x and y values
plot(df$x, df$y)

#fit a linear regression model to the data
reg_model <- lm(y ~ x, data = df)

#add the fitted regression line to the scatterplot
abline(reg_model)

```

Note that we simply need a value for the intercept and the slope to fit a simple linear regression line to the data using the abline() function. 

Thus, another way (although a more tedious way) of using abline() to add a regression line is to explicitly specify the intercept and slope coefficients of the regression model:

```{r}
#define dataset
df <- data.frame(x = c(1,1,2,3,4,4,5,6,7,7,8,9,10,10,11),
                 y = c(12,13,16,11,22,24,25,24,24,27,30,32,35,38,40))

#plot x and y values in dataset
plot(df$x, df$y, xlim = c(0,10), ylim = c(0,40))

#fit a linear regression model to the data
reg_model <- lm(y ~ x, data = df)
print(reg_model)
print(class(reg_model))
print(typeof(reg_model))

#define intercept and slope values
a <- coefficients(reg_model)[1] #intercept
b <- coefficients(reg_model)[2] #slope

print(a)
print(b)

#add the fitted regression line to the scatterplot
abline(a=a, b=b)

```

### Slope and intercept of the regression line
- Slope indicates the steepness of a line 
- Intercept indicates the location where it intersects an axis
- The slope and the intercept define the linear relationship between two variables, and can be used to estimate an average rate of change.
- The greater the magnitude of the slope, the steeper the line and the greater the rate of change.

```{r}

# y=2+5x
plot(1,type = "n", xlab = "The slope is positive 5. When x increases by 1, y increases by 5.The y-intercept is 2.",ylab = "",  xlim = c(0,4), ylim = c(0,20))

text(1,10,"y=2+5x")
abline(a=2, b=5)

# y=7.2-0.4x
plot(1,type = "n", xlab = "The slope is negative 0.4. When x increases by 1, y decreases by 0.4. The y-intercept is 7.2.",ylab = "",  xlim = c(0,4), ylim = c(0,8))

text(1,6,"y=7.2-0.4x")
abline(a=7.2, b=-0.4)

# y=-4
plot(1,type = "n", xlab = "The slope is 0. When x increases by 1, y neither increases/decreases. The y-intercept is -4.",ylab = "",  xlim = c(0,4), ylim = c(1,-5))

text(1,-3.5,"y=-4")
abline(a=-4, b=0, col="blue")

```
Usually, this relationship can be represented by the equation $y = b_0 + b_1x$, where $b_0$ is the $y-intercept$ and $b_1$ is the $slope$.<br><br>

#### Example
A company determines that job performance for employees in a production department can be predicted using the regression model $y = 130 + 4.3x$ <br>
where,

- $x$ is the hours of in-house training they receive (from 0 to 20)
- $y$ is their score on a job skills test.

- The value of the $y-intercept$ ($130$) indicates the average job skill score for an employee with no training.
- The value of the $slope$ ($4.3$) indicates that for each hour of training, the job skill score increases, on average, by $4.3$ points.


```{r}

# y=130+4.3x
plot(1,type = "n", xlab = "hours of in-house training",ylab = "score on a job skills test",  xlim = c(0,15), ylim = c(120,220))

title("Job Performance Prediction Model")

text(5,135,"y=130+4.3x")
abline(a=130, b=4.3, col="red")
grid (3,5, lty = 6, col = "cornsilk2")

```

### 3.4. Multiple Plot in Single Canvas

- `mfrow()`
- `mfcol()`

```{r}

# 100 random numbers 
x <- rnorm(100)
y <- rnorm(100)
z <- rpois(100,2)

```

#### 3.4.1. Multiple Plot: 2-rows & 1-column
```{r}
par(mfrow = c(2,1))
plot(x,y,pch=20)
plot(x,z,pch=20)
```

#### 3.4.2. Multiple Plot: 1-rows & 2-columns
```{r}

par(mfrow = c(1,2))
plot(x,y,pch=20)
plot(x,z,pch=20)

```

#### 3.4.3. Multiple Plot (row-wise): 2-rows & 2-columns
```{r}

par(mfrow = c(2,2))
plot(x,y,pch=20, main = "First")
plot(x,z,pch=20, main = "Second")
plot(y,x,pch=20, main = "Third")
plot(z,x,pch=20, main = "Forth")

```

#### 3.4.4. Multiple Plot (column-wise): 2-rows & 2-columns
```{r}

par(mfcol = c(2,2))
plot(x,y,pch=20, main = "First")
plot(x,z,pch=20, main = "Second")
plot(y,x,pch=20, main = "Third")
plot(z,x,pch=20, main = "Forth")

```

### 3.5. Points Variable

```{r}

# Plot of random numbers
x <- rnorm(100)
y <- x + rnorm(100)

plot(x,y)

# Factor Object containing Male and Female data

g <- gl(2,50, labels = c("Male","Female"))

print(g)
```

```{r}

# Blank Plot
plot(x,y, type = "n")

# Add Male points to graph
points(x[g=="Male"],y[g=="Male"], col = "light blue",pch=19)

# Add Female points to graph
points(x[g=="Female"],y[g=="Female"], col = "pink", pch=19)

```

### 4. 'grDevice' package

### 4.1. What is Graphic Device?
 
A graphics device is something where you can make a plot appear

- A window on your computer (screen device)
- A PDF file (file device)
- A PNG or JPEG file (file device)
- A scalable vector graphics (SVG) file (file device)

When you make a plot in R, it has to be "sent" to a specific graphics device.

1. The most common place for a plot to be "sent" is the **screen device**
    i) Screen device on **Mac** is launched with the `quartz()`
    ii) Screen device on **Windows** is launched with `windows()`
    iii) Screen device on **Unix/Linux** is launched with `x11()` <br>
    Note: There is only one screen device on a Platform (Windows, Mac, Unix/Linux).
2. For plots that may be printed out or be incorporated into a document (e.g. Papers/Reports, Slide Presentation), a **file device** is useful.
    i) `pdf()`

### 4.2. How to create Plot
I. Creating plot on **screen device**. <br>
II. Creating plot on **file device**.
  
#### I. Creating plot on 'screen device'
1. call a plotting Function 
    - `plot()` in **base**, e.g. `plot(x,y)`, 
    - `xyplot()` in **lattice**, or 
    - `qplot()` in **ggplot2** will default to sending a plot to the screen device
2. Plot appears on the **screen device**
3. Annotate plot if necessary
    - `title(main= "Old Faithful Geyser data")`

```{r}

library(datasets)

with(faithful,plot(eruptions,waiting)) ## Make plot appear on screen device
title(main= "Old Faithful Geyser data") ## Annotate plot

```

#### II. Creating plot on 'file device'
1. Explicitly launch a graphics device
    - `pdf(file="myplot.pdf")`
2. call a plotting Function (**Note:if you are using a file device, no plot will appear on the screen** )
    - `plot()` in **base** e.g. `plot(x,y)`, 
    - `xyplot()` in **lattice**, or 
    - `qplot()` in **ggplot2**
3. Annotate plot if necessary
    - `title(main= "Old Faithful Geyser data")`
4. Explicitly close graphics  device with `dev.off()` (**Note:This is very important step**)

```{r}

library(datasets)

pdf(file = "myplot.pdf") ## Open PDF file; create 'myplot.pdf' in working directory

with(faithful,plot(eruptions,waiting)) ## Create plot and send to a file (no plot appears on screen device)

title(main= "Old Faithful Geyser data") ## Annotate plot; still nothing on screen

dev.off() ## Close the PDF file device; Now the 'myplot.pdf' will appear on the screen.

```

### 4.3. Multiple Open Graphics Devices
It is possible to open multiple graphics devices (screen, file, or both).<br>
The **currently active graphics device** can be found by calling `dev.cur()` <br>
Every open graphics device is assigned an integer starting with 2 (there is no graphics device 1).<br>
You can change the active graphics device with `dev.set(<integer>)`

```{r}
# Open first device and plot data1
windows()
data1 <- 1:10
plot(data1,main="FIRST")
 
# Open second device and plot data2
windows()
data2 <- rnorm(10)
plot(data2, main="SECOND")


# 1. Currently active device `dev.cur()`
dev.cur()

# 2. Setting active device `dev.set(<integer>)`

## set active device and draw line1
dev.set(3)
abline(0,1, col = "red")

# 3. Switching active device `dev.prev()` & `dev.next()`

## Switch active device and draw line2
dev.set(dev.next())
abline(h = mean(data2), col = "red")

## Switch active device and draw legend
dev.set(dev.prev())
abline(0,1, col = "green", lwd=5)
legend("topleft",legend = "Data", pch = 1)

```

### 3.4. Copying Plots
The `dev.copy()` can be used to copy a plot from one device to another.

```{r}
library(datasets)
 
## Create plot on screen device
with(faithful, plot(eruptions, waiting))  
 
## Add a main title
title(main = "Old Faithful Geyser data")  
 
## Copy my plot to a PNG file
dev.copy(png, file = "geyserplot.png")  

## Don't forget to close the PNG device!
dev.off()  

```

### 5. Exploratory Graphs

Why do we use graphs in data analysis?
- To understand data properties
- To find patterns in data
- To suggest modeling strategies (use Linear/Non-linear Model)
- To 'debug' analysis (To see whats going wrong)
- To communicate results (To present result to peoples in the form of graphs)

Exploratory graphs are usually made very quickly and a lot of them are made in the process of checking out the data.

The goal of making exploratory graphs is usually developing a personal understanding of the data and to prioritize tasks for follow up.

### Case Study: Air Pollution in the United States

### Annual average PM2.5 averaged over the period 2008 through 2010

```{r}
pollution <- read.csv("data/avgpm25.csv", colClasses = c("numeric","character","factor","numeric","numeric"))
head(pollution)
```

#### Q. Do any states exceeds the standard of 12 mu g/m^3

* Simple Summaries of Data
  * One Dimension
  * Two Dimension and Beyond
  
#### I. One Dimension
1. Five Number Summary
2. Box plot
3. Histograms
4. Density Plot
5. Bar Plot

##### 1. Five Number Summary
```{r}
  summary(pollution$pm25)
```
##### 2. Box Plot
```{r}
  boxplot(pollution$pm25, col = "blue")

    # Overlaying Feature
  
    boxplot(pollution$pm25, col = "blue")
    abline(h=12)
  
    boxplot(pollution$pm25, col = "blue")
    abline(h=c(11,12,13))
```

##### 3. Histograms
```{r}  
  hist(pollution$pm25, col = "green")
  rug(pollution$pm25)
  
    # Histogram with change breaks
  
    hist(pollution$pm25, col = "green", breaks = 100)
    rug(pollution$pm25)
  
    # Overlaying Feature
  
    hist(pollution$pm25, col = "green")
    abline(v = 12,lwd = 2)
    abline(v = median(pollution$pm25), col = "magenta", lwd = 4)
```

##### 4. Bar plot
```{r} 
  barplot(table(pollution$region), col = "wheat", main = "Number of Counties in Each Region")
```

#### II. 2-Dimensional Graphs

```{r}
pollution <- read.csv("data/avgpm25.csv", colClasses = c("numeric","character","factor","numeric","numeric"))
head(pollution)
```

##### 1. Multiple Boxplots
```{r}
  boxplot(pm25 ~ region,data = pollution, col = "red")
```  

##### 2. Multiple Histograms
```{r}  
  par(mfrow = c(2,1),mar = c(4,4,2,1) )
  hist(subset(pollution, region == "east")$pm25, col = "green")
  hist(subset(pollution, region == "west")$pm25, col = "green")
```

##### 3. Scatterplot
```{r}  
  with(pollution, plot(latitude, pm25))
  abline(h = 12, lwd = 2, lty = 2)
  
  # 3.1 Scatterplot - Using Color
  with(pollution, plot(latitude, pm25, col = region))
  abline(h = 12, lwd = 2, lty = 2)
   
  # 3.2 Multiple Scatterplots
  par(mfrow = c(1,2),mar = c(5,4,2,1) )  
  with(subset(pollution, region == "west"), plot(latitude, pm25, main = "west"))
  with(subset(pollution, region =="east"), plot(latitude, pm25, main="East"))
```  